Mining of Keystroke and Mouse Dynamics to Increase the Engagement of Students with Programming Assignments

### 2018-04-19 paper.tex
Revises abstract please @mariosky take a look

% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Mining of Keystroke and Mouse Dynamics to Enhance Student Engagement with Programming Assignments}

\author{\authorname{Mario Garcia Valdez\sup{1}, Amaury Hernandez Aguila\sup{1},
  Juan-J. Merelo\sup{2} and Alejandra Mancilla Soto\sup{1}}
\affiliation{ Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico),  \email{\tt \(mario,amherag,alejandra.mancilla\)@tectijuana.edu.mx}}
\affiliation{Geneura Team and CITIC, University of Granada (Spain),
    \email{\tt jmerelo@ugr.es}}
}

\keywords{Affective computing, neural networks, machine learning}

\abstract{
When choosing from a collection of programming assignments, instructors must
gauge the student\'s affective state and skills to select a learning resource
with the appropriate level of difficulty. This work is
focused on capturing and pre-processing of data for the evaluation of several
machine learning techniques aim at classifing the affective
states of users with different levels of expertise
when learning a new programming language.
Currently, web-based platform is applied to collect data from the students\'
interactions whith programming excersises. The aim of this research is to
evaluate the use of keyboard and mouse dynamics as an appropriate
non-obtrusive sensory input for an affective aware system.
Feature vectors were obtained by pre-processing
data from keyboard and mouse dynamics captured from students as
they were coding basic Python programming assignments.
Feature vectors were used for the classification of five affective states:
boredom, frustration, distraction, relaxation and engagement.
The following classification algorithms have been evaluated: k -nearest
neighbors,feed forward neural networks, na\"ive Bayes classifier and the J-48
tree indiction algorithm). The best accuracy was around 75\% and was achieved
by the J-48 algorithm. Results show that data gathered from ready-available,
non-obtrusive sensors can be used successfully
as another input to hyebrid classification models in order to predict an
individual\'s affective states.}

>>>>

% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Mining of Keystroke and Mouse Dynamics to Increase the Engagement of Students with Programming Assignments}

\author{\authorname{Mario Garcia Valdez\sup{1}, Amaury Hernandez Aguila\sup{1},
  Juan-J. Merelo\sup{2} and Alejandra Mancilla Soto\sup{1}}
\affiliation{Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico),  \email{\tt \(mario,amherag,alejandra.mancilla\)@tectijuana.edu.mx}}
\affiliation{Geneura Team and CITIC, University of Granada (Spain),
    \email{\tt jmerelo@ugr.es}}
}

\keywords{Affective computing, neural networks, machine learning, learning analytics}

\abstract{The aim of this research is to 
evaluate the use of keyboard and mouse dynamics as an appropriate
non-obtrusive sensory input for an affective aware system, and is
motivated by the lack of tools and methodologies for taking into
account the affective state of the student in learning
environments. In an ideal situation, when instructors have to choose
from 
  a collection of programming assignments, they should consider the
  student\'s affective state and skills in order to select a learning resource
  with the appropriate level of difficulty. However, neither the data
  or the ability to process it are present in current learning
  management systems. This work tries to address this problem, by 
  focused on the capture and pre-processing of data that is going to
  be fed to several machine learning techniques with the objective of classifying the affective
states of users with different levels of expertise
when learning a new programming language. We capture student data from
a web-based platform, a learning management system where students
interact whith programming excersises. We introduce in this paper a
series of pre-processing techniques that are able to convert data from keyboard and mouse dynamics captured from students as
they were coding basic Python programming assignments  into feature
vector, that have been later used for the classification into five affective states:
boredom, frustration, distraction, relaxation and engagement.
The following classification algorithms have been evaluated: k -nearest
neighbors,feed forward neural networks, na\"ive Bayes classifier and the J-48
tree induction algorithm). The best accuracy was around 75\% and was achieved
by the J-48 algorithm. Results show that data gathered from ready-available,
non-obtrusive sensors can be used successfully
as another input to hyebrid classification models in order to predict an
individual\'s affective states, and these in turn can be used to make
students more engaged and thus learn more efficiently.}
